---
title: "Introduction to R"
subtitle: "Import and Clean Data in R"
output: html_document
---

## How to follow along and participate

```{r chunk1}
library(package = "here")
library(package = "tidyverse")
library(package = "haven")
```

## Importing data into R

* Before running this code, make sure the participant file and the data file are saved in the same folder on your computer

```{r chunk2}
# use here so that R looks in the same folder you are in
here()

# use read_spss function to read the spss file 
globalData2019 <- read_spss(file = "Pew Research Center Global Attitudes Spring 2019 Dataset WEB.sav")
```


## Selecting variables and filtering observations

```{r chunk3}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) 

```

## Examining the smaller data set

```{r chunk4}
summary(object = globalData2019clean)
```

## Changing data types 1

```{r chunk5}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female'))
```


## Checking your work 1

```{r chunk6}
summary(object = globalData2019clean)
```

## Changing data types 2

```{r chunk7}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                       '1' = 'Men',
                                       '2' = 'Women',
                                       '3' = 'Same/both equally',
                                       '8' = NA_character_,
                                       '9' = NA_character_))
```


## Checking your work 2

```{r chunk8}
summary(object = globalData2019clean)
```

## Changing data types 3

```{r chunk9}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_))
```



## Checking your work 3

```{r chunk10}
summary(object = globalData2019clean)
```



## Recoding data: You Try It!

* Look in the codebook to find the variable for **ECON_SIT** (note: it will be lower case in the codebook)

* Add recoding of **ECON_SIT** to the code below and run the code 

* Run the `summary()` code in the next code chunk to check your work

```{r chunk11}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) 
```

```{r chunk12}
# check your work
summary(object = globalData2019clean)
```

## Try Some More! 

* Recode the last two categorical variables, `BELIEVE_GOD` and `GLOBAL_COMMUNITY` to have appropriate category labels and `NA` 

* Use the codebook to find the labels you need 

* Add to the existing code: 

```{r chunk13}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_))
```


## Check your work: Try Some More!


```{r chunk14}
summary(object = globalData2019clean)
```

## Answer to Try Some More!

```{r try_more_key, include = FALSE}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_)) %>% 
  mutate(BELIEVE_GOD = recode_factor(BELIEVE_GOD,
                                     '1' = 'It is not necessary to believe in God in order to be moral and have good values',
                                     '2' = 'It is necessary to believe in God in order to be moral and have good values',
                                     '8' = NA_character_,
                                     '9' = NA_character_)) %>% 
  mutate(GLOBAL_COMMUNITY = recode_factor(GLOBAL_COMMUNITY,
                                          '1' = 'Part of a global community that works together to solve problems',
                                          '2' = 'Independent nations that compete with other countries and pursue their own interests',
                                          '3' = 'Both',
                                          '4' = 'Neither',
                                          '8' = NA_character_,
                                          '9' = NA_character_))
```

## Check your work: Try Some More!

```{r chunk15}
summary(object = globalData2019clean)
```


## The end

* Email me at: harrisj@wustl.edu

* Connect with me on Twitter: https://twitter.com/jenineharris
