---
title: "Introduction to R"
subtitle: "Create Basic Tables and Figures in R"
output: html_output
---


## How to follow along and participate

```{r chunk1}
library(package = "here")
library(package = "tidyverse")
library(package = "foreign")
library(package = "table1")
```


## Importing data into R

* Before running this code, make sure the participant file and the data file are saved in the same folder on your computer

```{r chunk2}
# use here so that R looks in the same folder you are in
here()

# use read_spss function to read the spss file 
globalData2019 <- read_spss(file = "Pew Research Center Global Attitudes Spring 2019 Dataset WEB.sav")
```



## Data cleaning (see prior video)

```{r chunk3}
globalData2019clean <- globalData2019 %>%
  filter(country == 21) %>% 
  select(SEX, AGE, BETTER_GENDER, LADDER_NOW, COUNTRY_SATIS, 
         ECON_SIT, BELIEVE_GOD, GLOBAL_COMMUNITY) %>% 
  zap_labels() %>% 
  mutate(SEX = recode_factor(SEX,
                             '1' = 'Male',
                             '2' = 'Female')) %>% 
  mutate(BETTER_GENDER = recode_factor(BETTER_GENDER,
                                    '1' = 'Men',
                                    '2' = 'Women',
                                    '3' = 'Same/both equally',
                                    '8' = NA_character_,
                                    '9' = NA_character_)) %>% 
  mutate(COUNTRY_SATIS = recode_factor(COUNTRY_SATIS,
                                       '1' = 'Satisfied',
                                       '2' = 'Dissatisfied',
                                       '8' = NA_character_,
                                       '9' = NA_character_)) %>% 
  mutate(ECON_SIT = recode_factor(ECON_SIT,
                                  '1' = 'Very good',
                                  '2' = 'Somewhat good',
                                  '3' = 'Somewhat bad',
                                  '4' = 'Very bad',
                                  '8' = NA_character_,
                                  '9' = NA_character_)) %>% 
  mutate(BELIEVE_GOD = recode_factor(BELIEVE_GOD,
                                     '1' = 'It is not necessary to believe in God in order to be moral and have good values',
                                     '2' = 'It is necessary to believe in God in order to be moral and have good values',
                                     '8' = NA_character_,
                                     '9' = NA_character_)) %>% 
  mutate(GLOBAL_COMMUNITY = recode_factor(GLOBAL_COMMUNITY,
                                          '1' = 'Part of a global community that works together to solve problems',
                                          '2' = 'Independent nations that compete with other countries and pursue their own interests',
                                          '3' = 'Both',
                                          '4' = 'Neither',
                                          '8' = NA_character_,
                                          '9' = NA_character_)) 
```


## Check your work

```{r chunk4}
summary(object = globalData2019clean)
```


## Basic table


```{r chunk5}
table1(~ SEX + AGE + BETTER_GENDER,
       data = globalData2019clean)
```


## Basic table with labels

* Add labels for the variables and add the `BELIEVE_GOD` variable

```{r chunk6}
label(globalData2019clean$SEX) <- "Sex"
label(globalData2019clean$AGE) <- "Age (years)" 
label(globalData2019clean$BETTER_GENDER) <- "Who has a better life in this country"
label(globalData2019clean$BELIEVE_GOD) <- "Which comes closest to your opinion"
```


## Basic table with labels

```{r chunk7}
table1(~ SEX + AGE + BETTER_GENDER + BELIEVE_GOD,
       data = globalData2019clean)
```


## Basic table with labels and grouped by sex

```{r chunk8}
table1(~ AGE + BETTER_GENDER + BELIEVE_GOD | SEX,
       data = globalData2019clean)
```

## Basic table with labels, groups, and total

```{r chunk9}
table1(~ AGE + BETTER_GENDER + BELIEVE_GOD | SEX,
       data = globalData2019clean, overall = "Total")
```


## You Try It!

* Copy and paste the label and table code from above into the empty R chunk

* Add labels to the `COUNTRY_SATIS` and `ECON_SIT` variables (check the codebook to find these)

* Add the `COUNTRY_SATIS` and `ECON_SIT` variables to the table 

* Run the code to make sure it works

```{r chunk10}

```


## Using the ggplot2 package for plotting

* Part of the `tidyverse` set of packages in R

* `gg` stands for **grammar** of **graphics**

* Plots created in `ggplot2` are built in layers 

* Let's look at some examples

* First, a bar chart...

## Basic bar chart

```{r chunk11}
globalData2019clean %>%
  ggplot(aes(x = SEX)) +
  geom_bar() 
```

## Fancy bar chart

```{r chunk12}
globalData2019clean %>%
  ggplot(aes(x = SEX)) +
  geom_bar(fill = "deeppink") +
  labs(x = "Sex",
       y = "Frequency",
       title = "Sex of 995 participants in a 2019 survey of people in Nigeria.") +
  theme_minimal()
```

## Fancy bar chart with colors

```{r chunk13}
globalData2019clean %>%
  ggplot(aes(x = SEX, fill = SEX)) +
  geom_bar() +
  labs(x = "Sex",
       y = "Frequency",
       title = "Sex of 995 participants in a 2019 survey of people in Nigeria.") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

## Fancy bar chart without legend

```{r chunk14}
globalData2019clean %>%
  ggplot(aes(x = SEX, fill = SEX)) +
  geom_bar() +
  labs(x = "Sex",
       y = "Frequency",
       title = "Sex of 995 participants in a 2019 survey of people in Nigeria.") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "none") 
```


## You try it!

Make a bar chart of the `BETTER_GENDER` responses from survey participants with the Figure 1 code in your R Markdown file ().

Edit the code to: 

* Use `BETTER_GENDER` as the fill variable in the `ggplot()` layer

* Change the color of the bars by replacing "Set2" with the name of another palette found on the <a href = "https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/">https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/</a> website

* Add labels for the axes and a title inside the quote marks (see codebook for more about the question)

* Replace `theme_minimal()` with another theme, you can find them here: https://ggplot2.tidyverse.org/reference/ggtheme.html

When you have completed the changes, click the little green arrow on the top right of the code chunk to run the code.

## Figure 1 code 

```{r chunk15}
# Fill in the Figure 1 code
globalData2019clean %>%
  ggplot(aes(x = BETTER_GENDER, fill = )) +
  geom_bar() +
  labs(x = "",
       y = "",
       title = "") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "none")
```


## Stacked bar plot

```{r chunk16}
globalData2019clean %>% 
  drop_na(BETTER_GENDER) %>% 
  ggplot(aes(x = BETTER_GENDER, fill = SEX)) +
  geom_bar() +
  labs(x = "Who has better life",
       y = "Frequency",
       title = "Who has a better life by sex according to 995 survey participants in Nigeria in 2019.") +
  scale_fill_brewer(palette = "BuPu") +
  theme_minimal() 
```

## Grouped bar plot

```{r chunk17}
# Fill in the Figure 1 code
globalData2019clean %>% 
  drop_na(BETTER_GENDER) %>% 
  ggplot(aes(x = BETTER_GENDER, fill = SEX)) +
  geom_bar(position = "dodge") +
  labs(x = "Who has better life",
       y = "Frequency",
       title = "Who has a better life by sex according to 995 survey participants in Nigeria in 2019.") +
  scale_fill_brewer(palette = "BuPu", name = "Survey participant sex") +
  theme_minimal() 
```


## You Try It!

* Create a grouped bar plot for Figure 2 with `BELIEVE_GOD` as the x-axis and `SEX` as the fill color 

* Drop the `NA` values for `BELIEVE_GOD`

* Add good labels for the x-axis, y-axis, title (may need the codebook for this)

* Format it using a palette you like  

* Fix the legend title so it is "Participant sex"

```{r chunk18}
# Figure 2 code goes here

```


## Answer to You Try It!

```{r chunk19}
# Figure 2 code goes here
globalData2019clean %>% 
  drop_na(BELIEVE_GOD) %>% 
  ggplot(aes(x = BELIEVE_GOD, fill = SEX)) +
  geom_bar(position = "dodge") +
  labs(x = "Which is closest to your opinion",
       y = "Frequency",
       title = "Belief in God by sex according to 995 survey participants in Nigeria in 2019.") +
  scale_fill_brewer(palette = "BuPu", name = "Survey participant sex") +
  theme_minimal() 

```

## Shorten long labels

```{r chunk20}
# Figure 2 code goes here
globalData2019clean %>% 
  drop_na(BELIEVE_GOD) %>% 
  ggplot(aes(x = BELIEVE_GOD, fill = SEX)) +
  geom_bar(position = "dodge") +
  labs(x = "Which is closest to your opinion",
       y = "Frequency",
       title = "Belief in God by sex according to 995 survey participants in Nigeria in 2019.") +
  scale_fill_brewer(palette = "BuPu", name = "Survey participant sex") +
  theme_minimal() +
  scale_x_discrete(labels = 
                     function(x) 
                       str_wrap(x, width = 30))
```


## Group percentages

```{r chunk21}
globalData2019clean %>% 
  group_by(BELIEVE_GOD, SEX) %>% 
  count() %>% 
  group_by(BELIEVE_GOD) %>% 
  mutate(perc = 100 * (n / sum(n))) 
```

## Grouped bar plot

```{r chunk22}
globalData2019clean %>% 
  group_by(BELIEVE_GOD, SEX) %>% 
  count() %>% 
  group_by(BELIEVE_GOD) %>% 
  mutate(perc = 100 * (n / sum(n))) %>% 
  drop_na(BELIEVE_GOD) %>% 
  ggplot(aes(x = BELIEVE_GOD, y = perc, fill = SEX)) +
  geom_col(position = "dodge") +
  labs(x = "Which captures your beliefs",
       y = "Percentage of belief group",
       title = "Belief in God by sex according to 995 survey participants in Nigeria in 2019.") +
  scale_fill_brewer(palette = "Paired", name = "Survey participant sex") +
  theme_minimal() +
  scale_x_discrete(labels = 
                     function(x) 
                       str_wrap(x, width = 30))
```

## Histogram

```{r chunk23}
# Histogram for age
globalData2019clean %>% 
  ggplot(aes(x = AGE)) +
  geom_histogram() 
```

## Fancy histogram

```{r chunk24}
# Histogram for age
globalData2019clean %>% 
  ggplot(aes(x = AGE, fill = SEX)) +
  geom_histogram(color = "white") +
  labs(x = "Participant age (years)",
       y = "Frequency",
       title = "Age distribution by sex for 995 survey participants in Nigeria in 2019.") +
  scale_fill_brewer(palette = "Set1") +
  facet_grid(rows = vars(SEX)) +
  theme_minimal() 
```

## Box plot

```{r chunk25}
# Fill in the Figure 1 code
globalData2019clean %>% 
  ggplot(aes(x = SEX, y = AGE)) +
  geom_boxplot() +
  labs(x = "Participant sex",
       y = "Age",
       title = "Age by sex for 995 survey participants in Nigeria in 2019.") +
  theme_minimal() 
```

## Fancier box plot

```{r chunk26}

globalData2019clean %>% 
  ggplot(aes(x = SEX, y = AGE, color = SEX)) +
  geom_violin() +
  geom_boxplot() +
  geom_jitter() +
  labs(x = "Participant sex",
       y = "Age",
       title = "Age by sex for 995 survey participants in Nigeria in 2019.") +
  theme_minimal() 
```

## Fancier box plot

```{r chunk27}
globalData2019clean %>% 
  ggplot(aes(x = SEX, y = AGE, color = SEX)) +
  geom_violin() +
  geom_boxplot(width = .4) +
  geom_jitter(alpha = .3) +
  labs(x = "Participant sex",
       y = "Age",
       title = "Age by sex for 995 survey participants in Nigeria in 2019.") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  coord_flip()
```



## The end

* Email me at: harrisj@wustl.edu

* Connect with me on Twitter: https://twitter.com/jenineharris


